$Z_3$ topological order in the quantum dimer-pentamer model

### 2016-08-05 qdpm_paper.tex
Worked on discussion and references

Over the last several decades, experimental and theoretical work has demonstrated that strongly interacting quantum many-body systems support exotic quantum phases of matter not described by the Landau symmetry breaking paradigm. One class of such quantum phases of matter are topologically ordered quantum liquids~\cite{Wen1990}, first realized experimentally via the fractional quantum Hall effect~\cite{FQHE????}. Despite the absence of conventional symmetry breaking and a local order parameter, topological quantum liquids possess a non-trivial quantum order that distinguishes them from trivial liquid phases~\cite{Nayak2008}. Although not detectable by local operators, topological order arises due to a non-local entanglement structure that leaves a signatures in the bipartite entanglement entropy~\cite{Levin2006a,Kitaev2006b}. Such topological phases may provide the basis for fault-tolerant quantum information processing~\cite{Freedman2001,Kitaev2003}.

Because of the exotic nature of topological phases and their potential use for quantum information processing, there is great interest in the development of theoretical models which display these phases. Such models offer the possibility of furthering the understanding of these phases as well as providing possibly candidate experimental systems in which to constructively engineer topological order~\cite{Duan2003,Jaksch2005,Lewenstein2007,Jiang2008d,Weimer2010a,Herdman2010c,Martinis2015}. A variety of exactly soluble lattice models are known to support ground states with topological order that ranges from the simplest variety, $Z_2$ topological order~\cite{Kitaev2003,Wen2003}, to those with non-Abelian anyonic excitations~\cite{Levin2005a,Kitaev2006a}; such models generally involve complex many-body interactions, and thus it is desirable to find simpler models supporting topological order that are closer to what is accessible experimentally.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t!]
    \centering
    \includegraphics[width=0.75\columnwidth]{QDPM_ex_config.pdf}
    \caption{An example of an allowed configuration in the quantum dimer-pentamer model, in which pentamers (four dimers touching at a vertex) are allowed.}
    \label{fig:QDPMex}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.8\linewidth]{example_local_constraints.pdf}
    \caption{Clockwise from the upper left panel we show an example of a QDM with constraints: one
    dimer ($n_0 = 1$), two dimers ($n_0 = 2$), and an even ($n_P = 0$) or odd ($n_P = 1$) number of dimers touching each vertex.}
    \label{fig:example_local_constraints}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Here we will provide background about generalized quantum dimer models on the square lattice. We consider systems with Ising-like dimer degrees of freedom on each link, such that each link has either $0$ or $1$ dimers. Given an Ising spin $\sigma_\ell$ on each link $\ell$ we define the number of dimers on the link $n_\ell \equiv 1/2+\sigma_\ell^z$, where $\sigma_\ell^i$ are the associated Pauli matrices. The Hilbert space consists of configurations of dimers on the square lattice that satisfy a local constraint at each vertex. We define a $U(1)$ charge $Q_v$ at each vertex $v$ of the lattice in terms of the number of dimers touching $v$, $n_v\equiv \sum_{\ell \in v}n_\ell$:
\begin{equation}
Q_v \equiv \pm \left( n_v - c_P \right)
\end{equation}
where $c_P = 2$, ($c_P=1$) for models in the even (odd) sector of the constraint, and the $+$($-$) sign corresponds to vertices on the $A$($B$) sublattice. A local dimer number constraint requires $Q_v$ to vanish for all vertices; a dimer parity constraint requires only $(Q_v\mod 2)$ to vanish.

@@

We define the string operator is defined as
\begin{equation}
S_\gamma \left( \alpha \right) \equiv e^{i \alpha \left( \Phi_\gamma - \tilde{\Phi}_\gamma \right)} \label{eq:Sgamma}
\end{equation}
where $\alpha$ such that $S_\gamma$ commutes with $\HGQDM$ everywhere except at the ends of $\gamma$. Therefore, for closed loops $\Gamma$, $[ S_\Gamma,\HGQDM]=0$. For open strings, $S_\gamma$ does not commute with the Hamiltonian and can be viewed as creating a pair of ``magnetic charges'' at the ends of the string: correspondingly $S_\gamma \ket{\Phi_{\RK}}$ can used as a trial wavefunction for an excited state with a pair of magnetic vortices. If a state has topological order with deconfined electric charges, these magnetic vortices should be gapped excitations, and correspondingly the magnetic string operator should decay exponentially~\cite{Read1989a,Senthil2000,Senthil2001e}. The condensation of magnetic vortices can drive a phase transition from a topologically ordered phase to a conventionally order phase~\cite{Jalabert1991,Ralko2007}. Thus, the magnetic string correlation can be used as test for deconfinement of electric charges.

@@

%%%%%%%%%%%%%%%%%%%%%%%
    \begin{figure}[]
        \centering
        \includegraphics[width=1.0\columnwidth]{monomer_cor_log.pdf}
        \caption{Monomer-monomer correlations function $\langle m_{\bm{0}} m_{\bm{r}} \rangle$ along the  in the QDPM RK ground state doped with a single pair of monomers on lattices up to \note{$L=48$}. The horizontal lines correspond to $L^{-4}$ with the color corresponding to the system size. \note{Owen: Do you have data for $L=64$? Can you make the subscripts in the y-axis label bold math? Maybe we should renomalize this plot by a factor of $L^2$.}}
        \label{fig:monomer_correlation}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%

@@

We have presented the quantum dimer-pentamer model on the square lattice and studied the ground state at the RK point. Using numerical calculations, we have demonstrated that the RK state of the QDPM is a dimer liquid, without spontaneously broken symmetry. The extensive topological degeneracy of the QDM has been reduced to a finite 9-fold degeneracy in the QDPM on a torus. We find that the dimer correlations decay exponentially with distance in contrast to the power-law decay of dimer correlations in the QDM. Additionally we also have demonstrated that monomer defects are deconfined in the QDPM, in contrast with the critical confinement of monomers in the QDM. Finally, we find that the $Z_3$ magnetic string correlations decay exponentially and display no evidence of $Z_3$ vortex condensation; this is consistent with $Z_3$ vortices being gapped quasiparticles excitations about the QDPM ground state.

The above results suggest a $Z_3$ gauge theory description of the low energy physics of the QDPM and the RK ground state possessing $Z_3$ topological order.

Future work could directly probe the existence of a finite gap above the RK point by studying the imaginary time dynamics of correlations functions using Monte Carlo~\cite{Henley}. Additionally, one could conclusively demonstrate the existence of $Z_3$ topological order by computing the topological entanglement entropy in the QDPM ground state~\cite{Levin2005,Kitaev2005,Hastings2010}.

A finite gap above the RK point suggests that a stable topologically ordered phase should exist surrounding the RK point; the neighboring phase diagram could be studied using quantum Monte Carlo methods.

Finally, given the stability of the topologically ordered phase, it may be possible to realize the QDPM in a more microscopically realistic spin model with two body interactions.

>>>>

Over the last several decades, experimental and theoretical work has demonstrated that strongly interacting quantum many-body systems support exotic quantum phases of matter not described by the Landau symmetry breaking paradigm. One class of such quantum phases of matter are topologically ordered quantum liquids~\cite{Wen1990}, first realized experimentally via the fractional quantum Hall effect~\cite{Tsui1982,Laughlin1983,Wen1990a,Wen1995}. Despite the absence of conventional symmetry breaking and a local order parameter, topological quantum liquids possess a non-trivial quantum order that distinguishes them from trivial liquid phases~\cite{Nayak2008}. Although not detectable by local operators, topological order arises due to a non-local entanglement structure that leaves a signatures in the bipartite entanglement entropy~\cite{Levin2006a,Kitaev2006b}. Such topological phases may provide the basis for fault-tolerant quantum information processing~\cite{Freedman2001,Kitaev2003}.

Because of the exotic nature of topological phases and their potential use for quantum information processing, there is great interest in the development of theoretical models which display these phases. Such models offer the possibility of furthering the understanding of these phases as well as providing candidate experimental systems in which to constructively engineer topological order~\cite{Duan2003,Jaksch2005,Lewenstein2007,Jiang2008d,Weimer2010a,Herdman2010c,Fowler2012a}. A variety of exactly soluble lattice models are known to support ground states with topological order that ranges from the simplest variety, $Z_2$ topological order~\cite{Kitaev2003,Wen2003}, to those with non-Abelian anyonic excitations~\cite{Levin2005a,Kitaev2006a}; such models generally involve complex many-body interactions, and thus it is desirable to find simpler models supporting topological order that are closer to what is accessible experimentally.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t!]
    \centering
    \includegraphics[width=0.75\columnwidth]{QDPM_ex_config.pdf}
    \caption{An example of an allowed configuration in the quantum dimer-pentamer model, in which pentamers (four dimers touching at a vertex) are allowed. \note{Owen: maybe we should stick with black and white for these cartoons}}
    \label{fig:QDPMex}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.8\linewidth]{example_local_constraints.pdf}
    \caption{Clockwise from the upper left panel we show an example of a QDM with constraints: one
    dimer ($n_0 = 1$), two dimers ($n_0 = 2$), and an even ($n_P = 0$) or odd ($n_P = 1$) number of dimers touching each vertex. \note{Owen: Maybe for consistency we should make all the dimers loos like the dimers in the upper-left.}}
    \label{fig:example_local_constraints}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Here we will provide background about generalized quantum dimer models on the square lattice. We consider systems with Ising-like dimer degrees of freedom on each link, such that each link has either $0$ or $1$ dimers. Given an Ising spin $\sigma_\ell$ on each link $\ell$ we define the number of dimers on the link $n_\ell \equiv 1/2+\sigma_\ell^z$, where $\sigma_\ell^i$ are the associated Pauli matrices. The Hilbert space consists of configurations of dimers on the square lattice that satisfy a local constraint at each vertex. We define a $U(1)$ charge $Q_v$ at each vertex $v$ of the lattice in terms of the number of dimers touching $v$, $n_v\equiv \sum_{\ell \in v}n_\ell$:
\begin{equation}
Q_v \equiv \pm \left( n_v - c_P \right)
\end{equation}
where $c_P = 2$, ($c_P=1$) for models in the even (odd) sector of the constraint, and the $+$($-$) sign corresponds to vertices on the $A$($B$) sublattice. A local dimer number constraint requires $Q_v$ to vanish for all vertices; a dimer parity constraint requires only $(Q_v\mod 2)$ to vanish.

@@

We define the string operator is defined as
\begin{equation}
S_\gamma \left( \alpha \right) \equiv e^{i \alpha \left( \Phi_\gamma - \tilde{\Phi}_\gamma \right)} \label{eq:Sgamma}
\end{equation}
where $\alpha$ such that $S_\gamma$ commutes with $\HGQDM$ everywhere except at the ends of $\gamma$. Therefore, for closed loops $\Gamma$, $[ S_\Gamma,\HGQDM]=0$. For open strings, $S_\gamma$ does not commute with the Hamiltonian and can be viewed as creating a pair of ``magnetic charges'' at the ends of the string: correspondingly $S_\gamma \ket{\Phi_{\RK}}$ can used as a trial wavefunction for an excited state with a pair of magnetic vortices. If a state has topological order with deconfined electric charges, these magnetic vortices should be gapped excitations, and correspondingly the magnetic string operator should decay exponentially~\cite{Read1989a,Senthil2000,Senthil2001e}. The condensation of magnetic vortices can drive a phase transition from a topologically ordered phase to a conventionally order phase~\cite{Jalabert1991,Ralko2007,Huh2011}. Thus, the magnetic string correlation can be used as test for deconfinement of electric charges.

@@

%%%%%%%%%%%%%%%%%%%%%%%
    \begin{figure}[]
        \centering
        \includegraphics[width=1.0\columnwidth]{monomer_cor_log.pdf}
        \caption{ The monomer-monomer correlations function $\langle m_{\bm{0}} m_{\bm{r}} \rangle$ along a coordinate axis in the QDPM RK ground state doped with a single pair of monomers on lattices up to \note{$L=48$}. The horizontal lines correspond to $L^{-4}$ for the system size corresponding to the color. \note{Owen: Do you have data for $L=64$? Can you make the subscripts in the y-axis label bold math? Maybe we should renomalize this plot by a factor of $L^2$. This is only one sub-lattice--what about the other sublattice?}}
        \label{fig:monomer_correlations}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%

@@

We have introduced the quantum dimer-pentamer model on the square lattice and studied the ground state at the RK point. Using numerical calculations, we have demonstrated that the RK state of the QDPM is a dimer liquid, without spontaneously broken symmetry. The extensive topological degeneracy of the square lattice QDM has been reduced to a finite 9-fold degeneracy in the QDPM on a torus. We find that the dimer correlations decay exponentially with distance in contrast to the power-law decay of dimer correlations in the QDM. Additionally we also have demonstrated that monomer defects are deconfined in the QDPM, in contrast with the critical confinement of monomers in the QDM. Finally, we find that the $Z_3$ magnetic string correlations decay exponentially and display no evidence of $Z_3$ vortex condensation; this is consistent with $Z_3$ vortices being gapped quasiparticles excitations above the QDPM ground state.

The above results suggest a the low energy physics of the QDPM is described by a $Z_3$ gauge theory with a $Z_3$ topological ordered ground state. Thus we have demonstrated that systems with Ising degrees of freedom and local constraints can lead to $Z_N$ topological order for $N>2$, providing the a new stable phase of matter beyond the commonly seen $Z_2$ topological spin liqud.

Future work could directly probe the existence of a finite gap above the RK point by studying the imaginary time dynamics of correlations functions using Monte Carlo methods similar to those used in here~\cite{Henley1997,Henley2004a}; such an approach could be used to explore if the low energy excitations are best described by $Z_3$ magnetic vorticies~\cite{Ivanov2004,Ralko2007,Misguich2008d}. Additionally, one could conclusively demonstrate the existence of $Z_3$ topological order by computing the topological entanglement entropy in the QDPM ground state Monte Carlo methods~\cite{Levin2005a,Kitaev2006b,Hastings2010}.

A finite gap above the RK point suggests that a stable topologically ordered phase should exist surrounding the RK point. While the Monte Carlo techniques used in this manuscript are limited to the RK point, the phase diagram near the RK point can be explored in future work using quantum Monte Carlo methods, as the QDPM has no sign problem. Away from RK, the condensation of $Z_3$ magnetic vortices could drive a continuous phase transition out of the topologically order phase, analogous to vison condensation in $Z_2$ spin liquids~\cite{Jalabert1991,Ralko2007,Huh2011,Hao2014,Slagle2014}; if such a continuous transition exists in the QDPM it may provide access to a new universality of class of quantum transitions~\cite{XU2012}.

Finally, given the stability of the topologically ordered phase, it may be possible to realize the QDPM in a more microscopically realistic spin model with two body interactions. The local constraints of the QDM can be realized in two-body spin-$1/2$ and Bose-Hubbard models, and the QDM Hamiltonian can arise perturbatively in the constrained low energy Hilbert space~\cite{Balents2002a,Zhitomirsky2005,Isakov2006c,Albuquerque2008}. Thus future work, it may be possible to engineer the QDPM in the low energy subspace of an experimentally realizable two-body model.
